### 2024-02-17 paper_sisc_first_revision/localpsf.tex
addressed most of Nicole's comments

We approximate the support of $\impulseresponse_x$ with the ellipsoid
%
%
\begin{equation}
\label{eq:support_ellipsoid}
E_x := \{x' \in \Omega: (x' - \spatialmean(x))^T \spatialcov(x)^{-1} (x' - \spatialmean(x)) \le \tau^2\},
\end{equation}
%
%
where $\tau$ is a fixed constant. The ellipsoid $E_x$ is the set of points within $\tau$ standard deviations of the mean of the Gaussian distribution with mean $\spatialmean(x)$ and covariance $\spatialcov(x)$, i.e., the Gaussian distribution which has the same mean and covariance as the normalized version of $\impulseresponse_x$. The quantity $\tau$ is a parameter that must be chosen appropriately. The larger $\tau$ is, the larger the ellipsoid $E_x$ is, and the more conservative the estimate is for the support of $\impulseresponse_x$. However, in Section~\ref{sec:sample_point_selection} we will see that the cost of the proposed method depends on how many non-overlapping ellipsoids $E_x$ we can ``pack'' in the domain $\Omega$ (more ellipsoids is better), and choosing a larger value of $\tau$ means that fewer ellipsoids will fit in $\Omega$. In practice, we find that $\tau=3.0$ yields a reasonable balance between these competing interests, and use $\tau=3.0$ in the numerical results. The fraction of the ``mass'' of $\impulseresponse_x$ residing outside of $E_x$ is less than $1/\tau^2$ by Chebyshev's inequality, though this bound is typically conservative.

@@

We use~\eqref{eq:vol_mean_var} to compute $\spatialvol$, $\spatialmean$, and $\spatialcov$ by applying $\Aop^T$ to polynomial functions. Then we use~\eqref{eq:support_ellipsoid} to form ellipsoid shaped estimates for the support of the $\phi_x$'s, \emph{without} computing them. This allows us to compute large numbers of $\impulseresponse_{x_i}$ in ``batches,'' $\eta_b$ (see Figure~\ref{fig:batches_intro}). We compute one batch, denoted $\eta_b$, by applying $\Aop$ to a weighted sum of point sources (Dirac comb) associated with a batch, $S_b$, of points $x_i$ scattered throughout $\Omega$ (Section~\ref{sec:get_impulse_response}). The batch of points, $S_b$, is chosen via a greedy ellipsoid packing algorithm so that, for $x_i,x_j \in S_b$, the support ellipsoid for $\impulseresponse_{x_i}$ and the support ellipsoid for $\impulseresponse_{x_j}$ do not overlap if $i \neq j$ (Section~\ref{sec:sample_point_selection}). Because these supports do not overlap (or do not overlap much), we can post process $\eta_b$ to recover the functions $\impulseresponse_{x_i}$ associated with all points $x_i \in S_b$. With one application of $\Aop$, we recover many $\impulseresponse_{x_i}$ (Section~\ref{sec:get_impulse_response}). The process is repeated until a desired number of batches is reached.

Once the batches $\eta_b$ are computed, we approximate the integral kernel $\Aker(y,x)$ at arbitrary points $(y,x)$ by interpolation of translated and scaled versions of the computed $\impulseresponse_{x_i}$ (Section~\ref{sec:approximate_kernel_entries}). The key idea behind the interpolation is the normalized local mean displacement invariance assumption discussed in Section~\ref{sec:local_mean_displacement_invariance}. Specifically, we approximate $\Aker(y,x) = \phi_x(y)$ by a weighted linear combination of the values $\frac{\spatialvol(x)}{\spatialvol(x_i)}\phi_{x_i}(y - \spatialmean(x) + \spatialmean(x_i))$ for a small number of sample points $x_i$ near $x$. The weights are determined by radial basis function (RBF) interpolation.

@@

We start with a finite set of candidate points $\candidatepts$ and build $\pointbatch_k$ incrementally with points selected from $\candidatepts$. For simplicity of explanation, here $\pointbatch_k$ and $\candidatepts$ are mutable sets that we add points to and remove points from. First we initialize $\pointbatch_k$ as an empty set. Then we select the candidate point $\candidatepoint_i \in \candidatepts$ that is the farthest away from all points in previous sample point batches $S_1 \cup S_2 \cup \dots \cup S_{k-1}$. Candidate points for the first batch $S_1$ are chosen randomly from $\candidatepts$.
Once $\candidatepoint_i$ is selected, we remove $\candidatepoint_i$ from $\candidatepts$. Then we perform the following checks:
\begin{enumerate}
	\item We check whether $\candidatepoint_i$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_{\candidatepoint_i} \cap E_{\candidatepoint_j} = \{\}$ for all $\candidatepoint_j \in \pointbatch_k$.
	\item We make sure that $\spatialvol(\candidatepoint_i)$ is not too small, by checking whether $\spatialvol(\candidatepoint_i) > \epsilon_V \spatialvol_\text{max}$. Here $\spatialvol_\text{max}$ is the largest value of $\spatialvol(\candidatepoint_j)$ over all points $q$ in the initial set of candidate points, and $\epsilon_V$ is a small threshold parameter (we use $\epsilon_V=10^{-5}$).
	\item We make sure that all eigenvalues of $\spatialcov(\candidatepoint_i)$ are positive, and the aspect ratio of $E_{\candidatepoint_i}$ (square root of the ratio of the largest eigenvalue of $\spatialcov(\candidatepoint_i)$ to the smallest) is bounded by a constant $1/\epsilon_\Sigma$ (we use $1/\epsilon_\Sigma=20$). Negative integral kernel entries due to discretization error can cause $\spatialcov(\candidatepoint_i)$ to be indefinite or highly ill-conditioned.
\end{enumerate}
If $\candidatepoint_i$ passes these checks 
(i.e., if $\candidatepoint_i$ is sufficiently far from other points in the batch, and $\spatialvol(\candidatepoint_i)$ and $\spatialcov(\candidatepoint_i)$ are acceptable)
then we add $\candidatepoint_i$ to $\pointbatch_k$. Otherwise we discard $\candidatepoint_i$. This process repeats until there are no more points in $\candidatepts$.  
We repeat the point selection process to construct several batches of points $\pointbatch_1, \pointbatch_2, \dots, \pointbatch_{\nbatch}$. For each batch, $\candidatepts$ is initialized as the set of all Lagrange nodes for the finite element basis functions used to discretize the problem, except for points in previous batches.

@@

We modify $\mathbf{A}_H$ to make it symmetric and remove negative eigenvalues via the following procedure. First, we symmetrize $\mathbf{A}_H$ via
$
\mathbf{A}_H^{\text{sym}} := \frac{1}{2}\left(\mathbf{A}_H + \mathbf{A}_H^T\right).
$
Next, we find negative eigenvalues and their corresponding eigenvectors for the generalized eigenvalue problem
$\mathbf{A}_H^{\text{sym}} \mathbf{u} = \lambda \mathbf{R}_H$
using a Cayley shift-and-invert Krylov scheme \cite{LehoucqEtAl98}. We flip the signs of these eigenvalues to be positive instead of negative (i.e., $\lambda \rightarrow |\lambda|$) by performing a low rank update to $\mathbf{A}_H^{\text{sym}}$. The primary computational task in the Cayley shift-and-invert scheme is the solution of shifted linear systems of the form
$\left(\mathbf{A}_H^{\text{sym}} + \mu_i \mathbf{R}_H\right) \mathbf{x} = \mathbf{b},$
for a small number of positive shifts $\mu_i$. We solve these linear systems by factorizing the matrices $\mathbf{A}_H^{\text{sym}} + \mu_i \mathbf{R}_H$ using fast H-matrix methods. We compute and flip all eigenvalues $\lambda < \epsilon_\text{flip}$ which are less than some threshold $\epsilon_\text{flip} \in (-1,0]$. By choosing $\epsilon_\text{flip} > -1$, we ensure that the modified version of $\mathbf{A}_H^{\text{sym}} + \mathbf{R}_H$ is positive definite. Choosing $\epsilon_\text{flip}=0$ would remove all erroneous negative eigenvalues. However, this is computationally infeasible if $\Aop$ has a large or infinite cluster of eigenvalues near zero, a common situation for Hessians in ill-posed inverse problems. We therefore recommend choosing $\epsilon_\text{flip} < 0$. In our numerical results, we use $\epsilon_\text{flip}=-0.1$.

@@

\paragraph{(1) Computing impulse response moments and batches} Computing $\spatialvol$, $\spatialmean$, and $\spatialcov$ requires applying $\Aop$ to $1$, $\gdim$, and $\gdim(\gdim+1)/2$ vectors, respectively.
Computing each $\eta_i$ requires applying $\Aop$ to one vector, so computing $\{\eta_i\}_{i=1}^{\nbatch}$ requires $\nbatch$ operator applications. In total, computing all impulse response moments and batches therefore requires
\begin{equation*}
	1 + \gdim + \gdim(\gdim+1)/2 + \nbatch \qquad \text{operator applications.}
\end{equation*}
In a typical application one might have $\gdim=2$ and $\nbatch=5$, in which case a modest $11$ operator applications are required.

>>>>

We approximate the support of $\impulseresponse_x$ with the ellipsoid
%
%
\begin{equation}
\label{eq:support_ellipsoid}
E_x := \{x' \in \Omega: (x' - \spatialmean(x))^T \spatialcov(x)^{-1} (x' - \spatialmean(x)) \le \tau^2\},
\end{equation}
%
%
where $\tau$ is a fixed constant (see Figure \ref{fig:frog_moments_ellipsoid}). The ellipsoid $E_x$ is the set of points within $\tau$ standard deviations of the mean of the Gaussian distribution with mean $\spatialmean(x)$ and covariance $\spatialcov(x)$, i.e., the Gaussian distribution which has the same mean and covariance as the normalized version of $\impulseresponse_x$. The quantity $\tau$ is a parameter that must be chosen appropriately. The larger $\tau$ is, the larger the ellipsoid $E_x$ is, and the more conservative the estimate is for the support of $\impulseresponse_x$. However, in Section~\ref{sec:sample_point_selection} we will see that the cost of the proposed method depends on how many non-overlapping ellipsoids $E_x$ we can ``pack'' in the domain $\Omega$ (more ellipsoids is better), and choosing a larger value of $\tau$ means that fewer ellipsoids will fit in $\Omega$. In practice, we find that $\tau=3.0$ yields a reasonable balance between these competing interests, and use $\tau=3.0$ in the numerical results. The fraction of the ``mass'' of $\impulseresponse_x$ residing outside of $E_x$ is less than $1/\tau^2$ by Chebyshev's inequality, though this bound is typically conservative.

@@

Here we provide an overview of the operator approximation algorithm. Then, in the subsections of this section (Sections \ref{sec:sample_point_selection}--\ref{sec:flip_eigs}) we present the technical details of the algorithm that were not already covered in Section \ref{sec:prerequisites}.

%\subsection{Algorithm overview}
%\label{sec:algorithm_overview}

We use~\eqref{eq:vol_mean_var} to compute $\spatialvol$, $\spatialmean$, and $\spatialcov$ by applying $\Aop^T$ to polynomial functions. Then we use~\eqref{eq:support_ellipsoid} to form ellipsoid shaped estimates for the support of the $\phi_x$'s, \emph{without} computing them (see Figure \ref{fig:frog_moments_ellipsoid}). This allows us to compute large numbers of $\impulseresponse_{x_i}$ in ``batches,'' $\eta_b$ (see Figures~\ref{fig:batches_intro}~and~\ref{fig:frog_batches}). We compute one batch, denoted $\eta_b$, by applying $\Aop$ to a weighted sum of point sources (Dirac comb) associated with a batch, $S_b$, of points $x_i$ scattered throughout $\Omega$ (Section~\ref{sec:get_impulse_response}). The batch of points, $S_b$, is chosen via a greedy ellipsoid packing algorithm so that, for $x_i,x_j \in S_b$, the support ellipsoid for $\impulseresponse_{x_i}$ and the support ellipsoid for $\impulseresponse_{x_j}$ do not overlap if $i \neq j$ (Section~\ref{sec:sample_point_selection}). Because these supports do not overlap (or do not overlap much), we can post process $\eta_b$ to recover the functions $\impulseresponse_{x_i}$ associated with all points $x_i \in S_b$. With one application of $\Aop$, we recover many $\impulseresponse_{x_i}$ (Section~\ref{sec:get_impulse_response}). The process is repeated until a desired number of batches is reached.

Once the batches $\eta_b$ are computed, we approximate the integral kernel $\Aker(y,x)$ at arbitrary points $(y,x)$ by interpolation of translated and scaled versions of the computed $\impulseresponse_{x_i}$ (Section~\ref{sec:approximate_kernel_entries} and Figure \ref{fig:hmatrix_neighbors_frog}). The key idea behind the interpolation is the normalized local mean displacement invariance assumption discussed in Section~\ref{sec:local_mean_displacement_invariance}. Specifically, we approximate $\Aker(y,x) = \phi_x(y)$ by a weighted linear combination of the values $\frac{\spatialvol(x)}{\spatialvol(x_i)}\phi_{x_i}(y - \spatialmean(x) + \spatialmean(x_i))$ for a small number of sample points $x_i$ near $x$. The weights are determined by radial basis function (RBF) interpolation.

@@

We start with a finite set of candidate points $\candidatepts$ and build $\pointbatch_k$ incrementally with points selected from $\candidatepts$. For simplicity of explanation, here $\pointbatch_k$ and $\candidatepts$ are mutable sets that we add points to and remove points from. First we initialize $\pointbatch_k$ as an empty set. Then we select the candidate point $\candidatepoint_i \in \candidatepts$ that is the farthest away from all points in previous sample point batches $S_1 \cup S_2 \cup \dots \cup S_{k-1}$. Candidate points for the first batch $S_1$ are chosen randomly from $\candidatepts$.
Once $\candidatepoint_i$ is selected, we remove $\candidatepoint_i$ from $\candidatepts$. Then we perform the following checks:
\begin{enumerate}
	\item We check whether $\candidatepoint_i$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_{\candidatepoint_i} \cap E_{\candidatepoint_j} = \{\}$ for all $\candidatepoint_j \in \pointbatch_k$.
	\item We make sure that $\spatialvol(\candidatepoint_i)$ is not too small, by checking whether $\spatialvol(\candidatepoint_i) > \epsilon_V \spatialvol_\text{max}$. Here $\spatialvol_\text{max}$ is the largest value of $\spatialvol(\candidatepoint_j)$ over all points $q$ in the initial set of candidate points, and $\epsilon_V$ is a small threshold parameter (we use $\epsilon_V=10^{-5}$).
	\item We make sure that all eigenvalues of $\spatialcov(\candidatepoint_i)$ are positive, and the aspect ratio of $E_{\candidatepoint_i}$ (square root of the ratio of the largest eigenvalue of $\spatialcov(\candidatepoint_i)$ to the smallest) is bounded by a constant $1/\epsilon_\Sigma$ (we use $1/\epsilon_\Sigma=20$). Negative integral kernel entries due to discretization error can cause $\spatialcov(\candidatepoint_i)$ to be indefinite or highly ill-conditioned.
\end{enumerate}
If $\candidatepoint_i$ passes these checks 
%(i.e., if $\candidatepoint_i$ is sufficiently far from other points in the batch, and $\spatialvol(\candidatepoint_i)$ and $\spatialcov(\candidatepoint_i)$ are acceptable)
then we add $\candidatepoint_i$ to $\pointbatch_k$. Otherwise we discard $\candidatepoint_i$. This process repeats until there are no more points in $\candidatepts$.  
We repeat the point selection process to construct several batches of points $\pointbatch_1, \pointbatch_2, \dots, \pointbatch_{\nbatch}$. For each batch, $\candidatepts$ is initialized as the set of all Lagrange nodes for the finite element basis functions used to discretize the problem, except for points in previous batches.

@@

We modify $\mathbf{A}_H$ to make it symmetric and remove negative eigenvalues via the following procedure. First, we symmetrize $\mathbf{A}_H$ via
$
\mathbf{A}_H^{\text{sym}} := \frac{1}{2}\left(\mathbf{A}_H + \mathbf{A}_H^T\right).
$
Next, we find negative eigenvalues and their corresponding eigenvectors for the generalized eigenvalue problem
$\mathbf{A}_H^{\text{sym}} \mathbf{u} = \lambda \mathbf{R}_H$
using a Cayley shift-and-invert Krylov scheme \cite{LehoucqEtAl98}. We flip the signs of these eigenvalues to be positive instead of negative (i.e., $\lambda \rightarrow |\lambda|$) by performing a low rank update to $\mathbf{A}_H^{\text{sym}}$. We observe that the eigenvectors associated with large erroneous negative eigenvalues tend to be directions that are nevertheless ``important'' to $\mathcal{A}$, so flipping the eigenvalues instead of setting them to zero tends to yield better approximations. The primary computational task in the Cayley shift-and-invert scheme is the solution of shifted linear systems of the form
$\left(\mathbf{A}_H^{\text{sym}} + \mu_i \mathbf{R}_H\right) \mathbf{x} = \mathbf{b},$
for a small number of positive shifts $\mu_i$. We solve these linear systems by factorizing the matrices $\mathbf{A}_H^{\text{sym}} + \mu_i \mathbf{R}_H$ using fast H-matrix methods. We compute and flip all eigenvalues $\lambda < \epsilon_\text{flip}$ which are less than some threshold $\epsilon_\text{flip} \in (-1,0]$. By choosing $\epsilon_\text{flip} > -1$, we ensure that the modified version of $\mathbf{A}_H^{\text{sym}} + \mathbf{R}_H$ is positive definite. Choosing $\epsilon_\text{flip}=0$ would remove all erroneous negative eigenvalues. However, this is computationally infeasible if $\Aop$ has a large or infinite cluster of eigenvalues near zero, a common situation for Hessians in ill-posed inverse problems. We therefore recommend choosing $\epsilon_\text{flip} < 0$. In our numerical results, we use $\epsilon_\text{flip}=-0.1$.

@@

\paragraph{(1) Computing impulse response moments and batches} Computing $\spatialvol$, $\spatialmean$, and $\spatialcov$ requires applying $\Aop^T$ to $1$, $\gdim$, and $\gdim(\gdim+1)/2$ vectors, respectively.
Computing each $\eta_i$ requires applying $\Aop$ to one vector, so computing $\{\eta_i\}_{i=1}^{\nbatch}$ requires $\nbatch$ operator applications. In total, computing all impulse response moments and batches therefore requires
\begin{equation*}
	1 + \gdim + \gdim(\gdim+1)/2 + \nbatch \qquad \text{operator applications.}
\end{equation*}
In a typical application one might have $\gdim=2$ and $\nbatch=5$, in which case a modest $11$ operator applications are required.
